Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-06-24 paper.tex
...

\author{Francisco Sant'Anna
        \\ Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ
        \\ francisco@ime.uerj.br}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus}
\caption{Pingus gameplay.
\label{fig.pingus}
}
\end{figure}

Pingus is an open-source puzzle-platform video game based on Lemmings.
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit (Figure~\ref{fig.pingus}).
% \footnote{Pingus gameplay video: \url{www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Pingus repository: \url{github.com/Pingus/pingus/}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot use long-lasting locals and loops, which are
elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \emph{goto}''.%
\footnote{``Callbacks as our Generations' \emph{goto} Statement'':
\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

In this work, we advocate structured synchronous reactive programming as a more
productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to the programming
language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
characteristics:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control-flow mechanisms, such
      as \code{spawn} and \code{await} (to create and suspend an activity).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      activity.
      There is no implicit preemption or real parallelism, resulting in
      deterministic execution.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection.

Our case study shows gains in productivity for six selected behaviors in the
game logic of Pingus through idiomatic code in \CEU.
We present an in-depth qualitative analysis of the proposed solutions in
comparison to the original implementations in C++.
We also identify four recurrent control-flow patterns that likely apply to
other games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
%
A control-flow pattern is a recurring technique to describe dependencies and
explicit orders between statements (or groups of statements) execution in a
program.
%For instance, consider how a key press stimulus propagates through the game
%entities and also what happens with them if the stimulus causes the end of the
%game.

\begin{figure*}[t]
\begin{verbatim}
Path              Ceu      C++   Ceu/C++    Descritpion
------------     ----     ----     ----     --------------------------------------------
game/            2064     2268     0.91     the main gameplay
  ./              710      679     1.05         main functionality
  objs/           470      478     0.98         world objects (tiles, traps, etc)
  pingu/          884     1111     0.80         pingu behaviors
    ./            343      458     0.75             main functionality
    actions/      541      653     0.83             pingu actions (bomber, climber, etc)
worldmap/         468      493     0.95     campaign worldmap
screens/         1109     1328     0.84     menus and screens
    option/       347      357     0.97         option menu
    others/       762      971     0.78         other menus and screens
misc/              56       46     1.22     miscellaneous functionality
                 ----     ----     ----
                 3697     4135     0.89
\end{verbatim}
\caption{The Pingus codebase directory tree.
\label{tab.tree}
}
\end{figure*}

The rewriting process consisted of identifying sets of callbacks implementing
control flow in the game and translating them to \CEU using appropriate
structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
We focus on a qualitative analysis for the programming techniques that we
applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%
We employed \emph{live code translation}, i.e., starting from the original
codebase in C++, we reimplemented piece-by-piece without breaking the game
compilation and execution.
%
This was only possible given the seamless integration between \CEU and
C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
happens at the source code level.
This enables trivial sharing of control and data, i.e., accessing C/C++ data
and calling C/C++ from \CEU and vice-versa.

@@

\section{The Pingus Codebase}
\label{sec.codebase}

In Pingus, the game logic accounts for almost half the size of the codebase:
18.173 from 39.362 LoCs (46\%) spread across 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
dealing with configurations and options, saved games and serialization, maps
and levels descriptions, string formatting, collision detection, graph
algorithms, etc.
This part remains unchanged and relies on the integration between \CEU and
C/C++.
%
Therefore, we only rewrote 9.186 LoCs spread across 126 files%
\footnote{Complete codebase: \url{github.com/an000/p/tree/master/cpp}}.
%
In order to only consider effective code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting 4.135 dense LoCs spread across 70 implementation files
originally written in C++%
\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
%
Figure~\ref{tab.tree} summarizes the effective game logic codebase in the two
implementations.

Although our analysis is mostly qualitative, the lines in Figure~\ref{tab.tree}
with lower ratio numbers correlate to the parts of the game logic that we
consider more susceptible to structured reactive programming.
For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
animations that are affected by timers, game rules, and user interaction.
In contrast, the \emph{Option screen} (\emph{ratio 0.97}) is a simple UI grid
with trivial mouse interactions.

@@

During the course of the rewriting process, we have identified four abstract
control-flow patterns which likely apply to other games as well:

@@

We describe six representative game behaviors in detail distributed in the four
patterns and analyse their implementations in C++ and \CEU.%
\footnote{Due to space constraints, we omit a fifth pattern
\emph{Signaling Mechanisms} and three other game behaviors.}

@@

Figure~\ref{lst.armageddon}.a shows the C++ implementation for the class
\code{ArmageddonButton} with methods for rendering the button and handling
mouse and timer events.
The code focus on the double click detection and hides unrelated parts with
\code{<...>}.
%
The methods \code{update} (ln. 14--26) and \code{on\_click} (ln. 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the base class
\code{RectComponent} (ln. 2), while the callback \code{update} continuously
reacts to the passage of time, frame by frame.
%Callbacks are short lived because they must react to input as fast as possible
%to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} (ln. 3) to track the
first click (ln. 32).
It also initializes the variable \code{press\_time} (ln. 4) to count the time
since the first click (ln. 16--17).
If another click occurs within 1 second, the class signals the double click to
the application (ln. 29--30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln. 18--21). 
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variables in the class, while the arrows 
represent the callbacks manipulating state.
%
Note in Figure~\ref{lst.armageddon} how the accesses to the state variables are
spread across the entire class.
For instance, the distance between the initialization of \code{pressed} (ln.
3) and the last access to it (ln. 32) is over 40 lines in the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}, which is defined in middle of the class (ln. 10--12), can
potentially access them.

@@

\CEU supports structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
In Figure~\ref{lst.armageddon}.b, the loop to detect double clicks (ln. 4--10)
awaits the first click (ln. 5) and then, while watching 1 second (ln. 6--9),
awaits the second click (ln. 7).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln. 8) and the \code{emit} in sequence signals the double click to the
application (ln. 12).
Otherwise, the \code{watching} block as a whole aborts after 1 second  and the
loop restarts, falling back to the first click \code{await} (ln. 5).
%
Double click detection in \CEU does not rely on state variables and is entirely
self-contained in the \code{loop} body.
Also, these 7 lines of code \emph{only} detect the double click, leaving the
actual effect to happen outside the loop (ln. 12), as well as all unrelated
code such as redrawing the button.

@@

In C++, the class \code{Bomber} in Figure~\ref{lst.bomber}.a defines the
callbacks \code{draw} and \code{update} to manage the state machine described
above.
%
The class first defines one state variable for each effect to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) first updates the pingu animation and
movement on every frame regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it switches to \emph{state-2} and
plays the ``Bomb!'' sound (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
41--46).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 34--37).
%
Note that a single numeric state variable suffices to track the states
(Figure~\ref{fig.bomber.fsm}), but the original developers probably chose to
encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., methods with mutable state).
%
Like the double click detection in C++, note that the state machine is encoded
across 3 different methods, each intermixing code with unrelated functionality
(e.g., changing frames, moving, and redrawing).

@@

Pingus supports 16 actions in the game.
As Figure~\ref{tab.actions} shows, 5 of them implement at least one state
machine and are considerable smaller in \CEU in terms of LoCs after eliminating
the state variables.
%
Considering the other 11 actions, the reduction in LoCs is negligible.
This asymmetry in the implementation of actions illustrates the gains in
expressiveness when describing state machines in direct style.

@@

Among all 65 implementation files in \CEU, we found 29 cases in 25 files using
structured mechanisms to substitute states machines.
They typically manifest as \code{await} statements in sequence.
%or in aborting constructs such as \code{par/or} and \code{watching}.

@@

\subsubsection{ Transition to the \emph{Credits Screen} from the \emph{Story Screen} }

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryDot::StoryDot(const FileReader& reader) :
  credits(false), // do not display by default
{
  <...>
  reader.read("credits", credits); // from file
}

void StoryDot::on_click() {
  <...>
  push_screen(<StoryScreen>(<...>, credits));
  <...>
}

@@

StoryScreenComp::StoryScreenComp (<...>) :
  credits(credits),
  <...>
{
  <...>
}

@@

void StoryScreenComp::next_text() {
  if (!displayed) {
    <...>
  } else {
    <...>
    if (!pages.empty()) {
      <...>
    } else {
      if (credits) {
        replace_screen(<Credits>(<...>));
      } else {
        pop_screen();
      }
    }
  }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
loop do
  var int ret = await Worldmap();
  if ret==STORY_CREDITS or ret==STORY_BACK then
    <...>
    var bool is_click = await Story();
    if is_click and ret==STORY_CREDITS then
      <...>
      await Credits();
    end
  else
    <...>
  end
end

@@

The campaign world map has clickable blue dots in the two extremes of the
progress trail for introductory and ending ambience stories, respectively.
For introductory stories, the game returns to the world map after displaying 
the story pages.
For ending stories, the game also displays a \emph{Credits screen} before
returning to the world map.%
\footnote{Credits screen animation: \url{github.com/an000/p/#4} }

In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
reads the level file (ln. 5) to check whether its an ending story and should,
after termination, display the \emph{Credits screen}.
%
The boolean variable \code{credits} is passed to the class
\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
to do after displaying the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
%
When the method \code{next\_text} has no pages left to display (ln. 32--38),
it decides where to go next, depending on the continuation flag
\code{credits} (ln. 33).

In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
the screens to display as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
and await its termination (ln. 5).
Finally, we check the returned values (ln. 6) to perhaps display the
\code{Credits} screen (ln. 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
class Bomber : public PinguAction {
    <...>
    Sprite sprite;
}

@@

void Bomber::draw (SceneContext& gc) {
    <...>
    gc.color().draw(sprite, <...>);
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName do
    <...>
    var Sprite sprite = spawn Sprite(<...>);
    <...>
end

@@

In contrast with C++, the screens in \CEU are completely decoupled and only the
\code{Main Loop} touches them: the \code{Worldmap} has no references to
\code{Story}, which has no references to \code{Credits}.

@@

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in five cases involving screen
transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay (e.g., \emph{walking} to \emph{jumping} and back to
\emph{walking}).

@@

In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
spawns a \code{Sprite} animation instance on its body (ln. 3).
%
The \code{Sprite} abstraction can react directly to external \code{update}
and \code{draw} events, bypassing the program hierarchy entirely.
Events in \CEU are broadcasted to the entire application in lexical order,
i.e., an abstraction that appears first in the source code (e.g., ln. 3) reacts
before another that appears second (e.g., ln. 4).
As discussed in Section~\ref{sec.pats.fsms.2}, this rule preserves determinism
and also respects the program static hierarchy.
While (\emph{and only while}) the bomber abstraction is alive, the sprite
animation remains alive.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.
Note that one can still declare a local event and restrict its visibility like
a local variable.

@@

\subsubsection{Case Study: Game UI Widgets}

@@

\subsubsection{Case Study: Managing the Pingus Lifecycle}
\label{sec.pats.lifespan.2}

@@

The implementation in \CEU has over 200 static instantiations spread across all
65 files.
For dynamic entities, it defines 23 pools in 10 files, with almost 96
instantiations across 37 files.
Pools are used to hold explosion particles, levels and level sets from files,
gameplay \& worldmap objects, and UI widgets.

@@

The author would like to thank Leonardo Kaplan and Alexander Tkachov for early
explorations and prototypes of the game rewrite.

>>>>

%\author{Francisco Sant'Anna
        %\\ Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ
        %\\ francisco@ime.uerj.br}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus-annotations}
\caption{Pingus gameplay.
\label{fig.pingus}
}
\end{figure}

Pingus is an open-source puzzle-platform video game based on Lemmings.
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit (Figure~\ref{fig.pingus}).
% \footnote{Pingus gameplay video: \url{www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Official Pingus repository: \url{github.com/Pingus/pingus/}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers), and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot use long-lasting locals and loops, which are
elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \emph{goto}''.%
\footnote{``Callbacks as our Generations' \emph{goto} Statement'':
\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

In this work, we advocate structured synchronous reactive programming as a more
productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to the programming
language \CEU~\cite{ceu.sensys13,ceu.mod15}.
\CEU has the following general characteristics:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item \emph{Reactive}: code executes in reactions to events.
\item \emph{Structured}: code uses structured control-flow mechanisms, such
      as \code{spawn} and \code{await} (to create and suspend an activity).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      activity.
      There is no implicit preemption or real parallelism, resulting in
      deterministic execution.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style in order to recover from the inversion of
control imposed by the observer pattern~\cite{rp.deprecating}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection for
memory management.

Our case study shows gains in productivity for six selected behaviors in the
game logic of Pingus rewritten in \CEU.
We present an in-depth qualitative analysis of the proposed solutions in
comparison to the original implementations in C++.
We also identify four recurrent control-flow patterns that likely apply to
other games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
%
A control-flow pattern is a recurring technique to describe execution
dependency and explicit ordering between statements (or groups of statements).
%For instance, consider how a key press stimulus propagates through the game
%entities and also what happens with them if the stimulus causes the end of the
%game.
We focus on a qualitative analysis for the programming techniques that we
applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%
We employed a \emph{live code rewrite}, i.e., starting from the original
codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
the game compilation and execution.

\begin{figure*}[t]
\begin{verbatim}
      Path              Ceu      C++   Ceu/C++    Descritpion
      ------------     ----     ----     ----     --------------------------------------------
 1    game/            2064     2268     0.91     the main gameplay
 2      ./              710      679     1.05         main functionality
 3      objs/           470      478     0.98         world objects (tiles, traps, etc)
 4      pingu/          884     1111     0.80         pingu behaviors
 5        ./            343      458     0.75             main functionality
 6        actions/      541      653     0.83             pingu actions (bomber, climber, etc)
 7    worldmap/         468      493     0.95     campaign worldmap
 8    screens/         1109     1328     0.84     menus and screens
 9        option/       347      357     0.97         option menu
10        others/       762      971     0.78         other menus and screens
11    misc/              56       46     1.22     miscellaneous functionality
                     ----     ----     ----
                     3697     4135     0.89
\end{verbatim}
\caption{The Pingus codebase directory tree.
\label{tab.tree}
}
\end{figure*}

@@

\section{The Pingus Codebase and Rewriting Process}
\label{sec.codebase}

% cd /opt/pingus/src
% sloccount .           -> 39975 -> (39362?)
% cd pingus/
% sloccount .           -> 18173

In Pingus, the game logic accounts for almost half the size of the codebase%
\footnote{
We used \emph{SLOCCount} to count only non-blank, non-comment lines in the
codebase: \url{www.dwheeler.com/sloccount/}
}:
18.173 from 39.362 LoCs (46\%) spread across 272 files.
%
However, about half of the game logic relates to non-reactive code, such as
dealing with configurations and options, saved games and serialization, maps
and level descriptions, string formatting, collision detection, graph
algorithms, etc.
This part remains unchanged and relies on the integration between \CEU and
C/C++.
%
Therefore, we only rewrote 9.186 LoCs spread across 126 files%
\footnote{\label{codebase} Effective codebase: \url{github.com/an000/p/tree/master/}}.
%
In order to only consider relevant code in the analysis, we then removed all
headers, declarations, trivial getters \& setters, and other innocuous
statements, resulting 4.135 dense LoCs spread across 70 implementation files
originally written in C++%
\footnotemark[\ref{codebase}].
%\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
\footnotemark[\ref{codebase}].
%\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
%
Figure~\ref{tab.tree} summarizes the effective game logic codebase in the two
implementations.

Although our analysis is mostly qualitative, the rows in Figure~\ref{tab.tree}
with lower ratio numbers correlate to the parts of the game logic that we
consider more susceptible to structured reactive programming.
For instance, the \emph{Pingu} behavior (row 4, \emph{ratio 0.80}) contains
complex animations that are affected by timers, game rules, and user
interaction.
In contrast, the \emph{Option screen} (row 9, \emph{ratio 0.97}) is a simple UI
grid with trivial mouse interactions.

The rewriting process consisted of identifying sets of callbacks implementing
control flow in the game and translating them to \CEU using appropriate
structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
This was only possible given the seamless integration between \CEU and
C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
happens at the source code level.
This enables trivial sharing of control and data, i.e., accessing C/C++ data
and calling C/C++ from \CEU and vice-versa.

@@

During the rewriting process, we have identified four abstract control-flow
patterns which likely apply to other games as well:

@@

We describe six representative game behaviors in detail distributed in the four
patterns and analyse their implementations in C++ and \CEU.%
\footnote{Due to space constraints, we omit three other game behaviors and a
fifth pattern \emph{Signaling Mechanisms}.}

@@

Figure~\ref{lst.armageddon}.a shows the C++ implementation for the class
\code{ArmageddonButton} with methods for rendering the button and handling
mouse and timer events.
The code in the figure focus on the double click detection and hides unrelated
parts with \code{<...>}.
%
The methods \code{update} (ln. 14--26) and \code{on\_click} (ln. 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the base class
\code{RectComponent} (ln. 2), while the callback \code{update} continuously
reacts to the passage of time, frame by frame.
%Callbacks are short lived because they must react to input as fast as possible
%to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} (ln. 3) to track the
first click (ln. 32).
It also initializes the variable \code{press\_time} (ln. 4) to count the time
since the first click (ln. 16--17).
If another click occurs within 1 second, the class signals the double click to
the application (ln. 29--30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln. 18--21). 
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variables in the class, and the arrows 
represent the callbacks manipulating state.
%
Note in Figure~\ref{lst.armageddon} how the accesses to the state variables are
spread across the entire class.
For instance, the distance between the initialization of \code{pressed} (ln.
3) and the last access to it (ln. 32) is over 40 lines in the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}, which is defined in middle of the class (ln. 10--12), can
potentially access them.

@@

\CEU supports structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
In Figure~\ref{lst.armageddon}.b, the loop to detect double clicks (ln. 4--10)
awaits the first click (ln. 5) and then, while watching 1 second (ln. 6--9),
awaits the second click (ln. 7).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln. 8) and the \code{emit} in sequence signals the double click to the
application (ln. 12).
Otherwise, the \code{watching} block as a whole aborts after 1 second  and the
loop restarts, falling back to the first click \code{await} (ln. 5).
%
Double click detection in \CEU does not rely on state variables and is entirely
self-contained in the \code{loop} body.
Also, those 7 lines of code \emph{only} detect the double click, leaving the
actual effect to happen outside the loop (ln. 12), as well as all unrelated
code such as redrawing the button.

@@

In C++, the class \code{Bomber} in Figure~\ref{lst.bomber}.a defines the
callbacks \code{draw} and \code{update} to manage the state machine of
Figure~\ref{fig.bomber.fsm}.
%
The class first defines one state variable for each effect to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) first updates the pingu animation and
movement on every frame, regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it switches to \emph{state-2} and
plays the ``Bomb!'' sound (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
41--46).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 34--37).
%
Note that a single numeric state variable would suffice to track the states as
in Figure~\ref{fig.bomber.fsm}, but the original developers probably chose to
encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., methods with mutable state).
%
Like the double click detection in C++, note that the state machine is encoded
across 3 different methods, each intermixing code with unrelated functionality
(e.g., changing frames, moving, and redrawing).

@@

Pingus supports 16 actions in the game.
5 of them, enumerated in Figure~\ref{tab.actions}, implement at least one state
machine and are considerable smaller in \CEU in terms of LoCs (after
eliminating the state variables).
%
For the other 11 actions without state machines, the reduction in LoCs is
negligible.
%
This asymmetry illustrates the gains in expressiveness when describing state
machines in direct style.

@@

Among all 65 implementation files in \CEU, we found 29 cases in 25 files that
use structured mechanisms to substitute states machines.
They typically manifest as \code{await} statements in sequence.
%or in aborting constructs such as \code{par/or} and \code{watching}.

@@

\subsubsection{ Transition from the \emph{Story Screen}  to the \emph{Credits Screen}}

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryDot::StoryDot(const FileReader& reader) :
  show_credits(false), // do not show by default
{
  <...>
  reader.read("credits", show_credits);
}             // from file

void StoryDot::on_click() {
  <...>
  push_screen(<StoryScreen>(show_credits));
  <...>
}

@@

StoryScreenComp::StoryScreenComp (<...>) :
  show_credits(show_credits),
  <...>
{
  <...>
}

@@

void StoryScreenComp::next_text() {
  if (!displayed) {
    <...>
  } else {
    <...>
    if (!pages.empty()) {
      <...>
    } else {
      if (show_credits) {
        replace_screen(<Credits>(<...>));
      } else {
        pop_screen();
      }
    }
  }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
loop do
  var int ret = await Worldmap();
  if ret==STORY_CREDITS or ret==STORY_BACK then
    <...>
    var bool is_click = await Story();
    if is_click and ret==STORY_CREDITS then
      <...>
      await Credits();
    end
  else
    <...>
  end
end

@@

The campaign world map has clickable blue dots in the two extremes of the
progress trail for introductory and ending ambience stories, respectively.
For introductory stories, the game returns to the world map after showing 
the story pages.
For ending stories, the game also shows a \emph{Credits screen} before
returning to the world map.%
\footnote{Credits screen animation: \url{github.com/an000/p/#4} }

In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
reads the level file (ln. 5) to check whether its an ending story and should,
after termination, show the \emph{Credits screen}.
%
The boolean variable \code{credits} is passed to the class
\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
to do after showing the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
%
When the method \code{next\_text} has no pages left to display (ln. 32--38),
it decides where to go next, depending on the continuation flag
\code{credits} (ln. 33).

In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
the screens to show as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
and await its termination (ln. 5).
After showing the story, we check the returned values (ln. 6) to perhaps show
the \code{Credits} screen (ln. 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
class Bomber : public Action {
    <...>
    Sprite sprite;
}

@@

void Bomber::draw () {
    <...>
    sprite.draw();
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName do
    <...>
    var Sprite sprite = spawn Sprite(<...>);
    <...>
end

@@

In contrast with C++, the screens in \CEU are decoupled and only the
\code{Main Loop} touches them: the \code{Worldmap} has no references to
\code{Story}, which has no references to \code{Credits}.

@@

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in five cases involving screen
transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay (e.g., \emph{walking} to \emph{falling} and back to
\emph{walking}).

@@

In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
spawns a \code{Sprite} animation instance on its body (ln. 3).
%
The \code{Sprite} abstraction can react directly to external \code{update}
and \code{draw} events, bypassing the program hierarchy entirely.
Events in \CEU are broadcasted to the entire application in lexical order,
i.e., an abstraction that appears first in the source code (e.g., ln. 3) reacts
before another one that appears second (e.g., ln. 4).
As discussed in Section~\ref{sec.pats.fsms.2}, this rule preserves determinism
and also respects the program static hierarchy.
While (\emph{and only while}) the bomber abstraction is alive, the sprite
animation remains alive.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.
Note that one can still declare a local event to restrict its visibility like
a local variable.

@@

\subsubsection{Case Study: Static Game UI Widgets}

@@

\subsubsection{Case Study: Dynamic Pingus Lifecycle}
\label{sec.pats.lifespan.2}

@@

The implementation in \CEU has over 200 static instantiations spread across all
65 files.
For dynamic entities, it defines 23 pools in 10 files, with almost 96
instantiations across 37 files.
Pools are used to hold explosion particles, levels and level sets from files,
gameplay \& worldmap objects, and also UI widgets.

@@

The authors would like to thank Leonardo Kaplan and Alexander Tkachov for early
explorations and prototypes of the game rewrite.
